---
title: "BIOSTAT 650 SEC 002 HW4"
author: "Jaehoon Kim"
date: "2024-09-30"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("sas7bdat")
library("corrplot")
library("gtsummary")
library("dplyr")
library("ggplot2")
```
\newcommand{\sumall}{\sum_{i=1}^n}
\newcommand{\EX}{\mathbb{E}}
\newcommand{\matr}[1]{\mathbf{#1}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\cov}{\operatorname{Cov}}

```{r chunk2}
mySASData = read.sas7bdat("C:/Users/aquil/Desktop/STAT 650/completedata.sas7bdat")

df <- mySASData
reg <- lm(Depression ~ Fatalism+Age+Sex+R_E, df)
summary(reg)

#define new observation
avg_fatal <- mean(df$Fatalism)
newdata = data.frame(Fatalism=avg_fatal, Age=70, Sex=1, R_E=1)

#use model to predict points value
predict(reg, newdata, se.fit=TRUE)

#estimated sigma squared
summary(reg)$sigma
```
```{R Variance}

X = df[c("Fatalism", "Age", "Sex", "R_E")]
X = cbind(1,X)
X_i = cbind(a=1, newdata)


VarYhat <- 
  (summary(reg)$sigma)^2*as.matrix(X_i)%*%solve(t(as.matrix(X))%*%as.matrix(X))%*%t(as.matrix(X_i))

VarYhat
sqrt(VarYhat)



```
## Problem 1.
Adjusting for demographic variables age, sex, and race/ethnicity, we estimate on average there is a 0.25 unit points increase in depression for every unit point increase in fatalism. This association is statistically significant with a p-value 1.08e-10 at a significance level $\alpha = 0.05.$

## Problem 2.
The adjusted $R^2$ of 0.1006 indicates that approximately 10.06% of the variation of the depression variable is explained by the multiple linear regression model containing fatalism, age, sex, race/ethnicity variables. The model shows a slight improvement as a predictor for the depression score compared to the simple model with only the fatalism variable.

## Problem 3.
The fitted model prediction estimates that on average 70 year old, female, Mexican Americans have a depression score of approximately 5.611. The variance of the estimate is $$\var(\hat{Y}_i) = \var(\mathbf{X}_i^T\hat{\boldsymbol{\beta}}) = \mathbf{X}_i^T \var(\hat{\boldsymbol{\beta}}) \mathbf{X}_i = \hat{\sigma}^2\mathbf{X}_i^T (\mathbf{X}^T\mathbf{X})^{-1} \mathbf{X}_i=0.127835$$ Thus, the standard error of the estimate is approximately $\sqrt{\var(\hat{Y}_i)} = 0.3575$.

## Problem 4.

(a)

All the matrices $\mathbf{A}_1, \mathbf{A}_2, \mathbf{A}_3$ are symmetrical and idempotent. We prove idempotency as such:

$$
\begin{aligned}
\mathbf{A}_1^2 = \mathbf{A}_1\mathbf{A}_1
&=\frac{1}{9}\begin{bmatrix}
1 &1 &1\\
1 &1 &1\\
1 &1 &1\\
\end{bmatrix}
\begin{bmatrix}
1 &1 &1\\
1 &1 &1\\
1 &1 &1\\
\end{bmatrix}\\
&=\frac{1}{9}\begin{bmatrix}
3 &3 &3\\
3 &3 &3\\
3 &3 &3\\
\end{bmatrix}
=\frac{1}{3}\begin{bmatrix}
1 &1 &1\\
1 &1 &1\\
1 &1 &1\\
\end{bmatrix}
\end{aligned}
$$

$$
\begin{aligned}
\mathbf{A}_2^2 = \mathbf{A}_2\mathbf{A}_2
&=\frac{1}{4}\begin{bmatrix}
1 &-1 &0\\
-1 &1 &0\\
0 &0 &0\\
\end{bmatrix}
\begin{bmatrix}
1 &-1 &0\\
-1 &1 &0\\
0 &0 &0\\
\end{bmatrix}\\
&=\frac{1}{4}\begin{bmatrix}
2 &-2 &0\\
-2 &2 &0\\
0 &0 &0\\
\end{bmatrix}
=\frac{1}{2}\begin{bmatrix}
1 &-1 &0\\
-1 &1 &0\\
0 &0 &0\\
\end{bmatrix}
\end{aligned}
$$

$$
\begin{aligned}
\mathbf{A}_3^2 = \mathbf{A}_3\mathbf{A}_3
&=\frac{1}{36}\begin{bmatrix}
1 &1 &-2\\
1 &1 &-2\\
-2 &-2 &4\\
\end{bmatrix}
\begin{bmatrix}
1 &1 &-2\\
1 &1 &-2\\
-2 &-2 &4\\
\end{bmatrix}\\
&=\frac{1}{36}\begin{bmatrix}
6 &6 &-4\\
6 &6 &-4\\
-4 &-4 &0\\
\end{bmatrix}
=\frac{1}{6}\begin{bmatrix}
1 &1 &-2\\
1 &1 &-2\\
-2 &-2 &4\\
\end{bmatrix}
\end{aligned}
$$

$\mathbf{A}_1$ has rank 1, $\mathbf{A}_2$ has rank 1, $\mathbf{A}_3$ has rank 1. Accordingly, given the quadratic form $Q_i=\mathbf{Y}^T\mathbf{A}_i\mathbf{Y}$ where $\mathbf{Y} \sim N(\boldsymbol{\mu}, \sigma^2\mathbf{I})$ for $i = 1, 2, 3,$

$$\frac{Q_i}{\sigma^2}=\frac{\mathbf{Y}^T\mathbf{A}_i\mathbf{Y}}{\sigma^2} \sim \chi_{_{1,\lambda}}$$

with non-centrality parameter $\lambda=\frac{\mu^T\mathbf{A}_i\mu}{\sigma^2}, \space \forall i = 1, 2, 3$

(b)

$$\mathbf{A}_1\mathbf{A}_2=\frac{1}{12}\begin{bmatrix}
2-2 &-2+2 &0\\
2-2 &-2+2 &0\\
2-2 &-2+2 &0\\
\end{bmatrix}=\mathbf{0}_{3 \times 3}$$

$$\mathbf{A}_1\mathbf{A}_3=\frac{1}{18}\begin{bmatrix}
2-2 &-2+2 &-4+4\\
2-2 &2-2 &-4+4\\
2-2 &-2+2 &-4+4\\
\end{bmatrix}=\mathbf{0}_{3 \times 3}$$

$$\mathbf{A}_2\mathbf{A}_3=\frac{1}{12}\begin{bmatrix}
1-1 &1-1 &-2+2\\
2-2 &2-2 &-4+4\\
2-2 &2-2 &-4+4\\
\end{bmatrix}=\mathbf{0}_{3 \times 3}$$

Thus, the $Q_i = \mathbf{Y}^T\mathbf{A}_i\mathbf{Y}$ are pairwise independent.